Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-13 paper/ResponseToReviews.tex

- As the variability of the viral RNA load appears to be the central result of the manuscript, it would be useful to:

    a. Employ a statistic to quantify the variability (e.g. entropy or Gini index)

    b. Use simple models to illustrate how much variability can be accounted for by simple effects, such as expected Poisson co-infection frequency, or the likelihood of attaining full complement of RNP genes.

{\color{black}
The idea of quantifying the variability in viral mRNA across infected cells is a good one.
We have done this using the Gini coefficient.
The results are in Figure 4--Figure supplement 4, and are referred to in the text related to that figure.
As a fun point of comparison, the Gini coefficients for the variability of viral mRNA across cells ($\ge$0.64 for all samples) exceeds the Gini coefficient for the unevenness of income distribution in the United States.
}

{\color{red}
Still need to think about a simple model.
}

### 2018-01-13 paper/paper.tex
added link to raw data on GEO
added Gini coefficients to quantify unevenness

\centerline{\includegraphics[width=0.85\linewidth]{figures/p_frac_flu_summary.pdf}}

@@

@@

\figsupp[Fraction of total viral mRNA derived from a given fraction of infected cells.\label{figsupp:cumulfracflu}]{

@@

For instance, the plot for the 8-hour sample shows that roughly 50\% of all viral mRNA is derived from 10\% of the infected cells.

@@

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
Notably, Figure~\ref{fig:viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).

@@

All time points except for the second 8-hour sample (8hr-2) were prepared on the same day.
For the infections, A549 cells were seeded in a 6-well plate, with two wells per time point. 
A single well of cells was trypsinized and counted prior to initiation of the experiment for the purposes of calculating MOI.
Wild-type and synonymously barcoded virus were mixed to an estimated ratio of 1:1 based on prior, exploratory, single-cell analyses (data not shown). 
At the initiation of our experiment, the wells for all time points were changed from D10 to influenza growth media.
Cells were then infected with 0.3 HA-expressing units of virus per cell (a determined by flow cytometry).
The infections were performed in order of time point: first the 10-hour time point, then the 8-hour, and then the 6-hour time point.
At one hour after infection, the media for each time point was changed to fresh influenza growth media.
Note that the HA-expressing units were calculated without this additional washing step, and so likely represent an overestimate of our final infectious dose (consistent with the fact that fewer than 30\% of cells appear infected in the single-cell sequencing data).
All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media .
For 8hr-2 sample, cells were infected as above except that the cells were infected at 0.1 HA-expressing units of virus per cell but no wash step was performed, and the sample was prepared on a different day.
After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the manufacturer's provided by 10x Genomics for the Chromium Single Cell platform.
All subsequent steps through library preparation followed the manufacturer's protocol.
Samples were sequenced on an Illumina HiSeq. 
The deep sequencing data are available on the Sequence Read Archive under accession \jdbcomment{Upload in progress.}

@@

Briefly, the raw deep sequencing data were processed using the 10X Genomics software package CellRanger (version 2.0.0).
The reads were aligned to a concatenation of the human and influenza virus transcriptomes.
The human transcriptome was generated by filtering genome assembly GRCh38 for protein coding genes defined in the GTF file GRCh38.87.
The influenza virus transcriptome was generated from the reverse-complement of the wildtype WSN viral RNA sequences as encoded in the reverse-genetics plasmids (Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}).
CellRanger calls cells based on the number of observed cell barcodes, and creates a cell-gene matrix.
We used custom Python code to annotate the cells in this matrix by the number of viral reads that could be assigned to the wildtype and synonymously barcoded virus.
Only about half of the viral reads overlapped the barcoded regions of the genes (Figure~\ref{fig:workflow}A) and could therefore be assigned to a viral barcode (Figure~\ref{fig:viralbarcodes}-Figures~supplement~\ref{figsupp:barcodescalled}).
So for calculations of the number of reads in a cell derived from each viral barcode for each viral gene, the total number of detected molecules of that gene are multiplied by the fraction of those molecules with assignable barcodes that are assigned to that barcode.
This annotated cell-gene matrix is in Supplementary~file~\ref{suppfile:cellgenematrix}.
A Jupyter notebook that performs these analyses is in Supplementary~file~\ref{suppfile:code}.

>>>>

\centerline{\includegraphics[width=0.83\linewidth]{figures/p_frac_flu_summary.pdf}}

@@

Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu} shows these same data in a cumulative fraction plots and calculates Gini coefficients to quantify the heterogeneity in viral mRNA load.

@@

\figsupp[Fraction of total viral mRNA from a given fraction of infected cells, and associated Gini coefficients.\label{figsupp:cumulfracflu}]{

@@

For instance, the plot for the 8hrs sample shows that $\sim$50\% of all viral mRNA is derived from $\sim$8\% of the infected cells.
The facet titles above each plot also give the Gini coefficient~\citep{gini1921measurement} that calculates the heterogeneity in the distribution of viral mRNA among infected cells.
Gini coefficients of 0 indicate a perfectly even distribution across cells, and Gini coefficients of 1 indicate a maximally skewed distribution.

@@

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
One way to quantify the heterogeneity of a distribution is to calculate the Gini coefficient~\cite{gini1921measurement}, which ranges from 0 for a completely uniform distribution, to 1 for a maximally skewed distribution.
Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu} shows the Gini coefficients for the distribution of viral mRNA across infected cells for each sample.
The Gini coefficients are $\ge$0.64 for all samples.
As a fun point of comparison, these Gini coefficients indicate that the distribution of viral mRNA across infected cells is more uneven than the distribution of income in the United States~\citep{alvaredo2011note}.

Notably, Figure~\ref{fig:viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).
This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).

@@

All time points except for the second 8-hour sample (8hr-2) were prepared on the same day.
For the infections, A549 cells were seeded in a 6-well plate, with two wells per time point. 
A single well of cells was trypsinized and counted prior to initiation of the experiment for the purposes of calculating MOI.
Wild-type and synonymously barcoded virus were mixed to an estimated ratio of 1:1 based on prior, exploratory, single-cell analyses (data not shown). 
At the initiation of our experiment, the wells for all time points were changed from D10 to influenza growth media.
Cells were then infected with 0.3 HA-expressing units of virus per cell (a determined by flow cytometry).
The infections were performed in order of time point: first the 10-hour time point, then the 8-hour, and then the 6-hour time point.
At one hour after infection, the media for each time point was changed to fresh influenza growth media.
Note that the HA-expressing units were calculated without this additional washing step, and so likely represent an overestimate of our final infectious dose (consistent with the fact that fewer than 30\% of cells appear infected in the single-cell sequencing data).
All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media .
For 8hr-2 sample, cells were infected as above except that the cells were infected at 0.1 HA-expressing units of virus per cell but no wash step was performed, and the sample was prepared on a different day.
After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the manufacturer's provided by 10x Genomics for the Chromium Single Cell platform.
All subsequent steps through library preparation followed the manufacturer's protocol.
Samples were sequenced on an Illumina HiSeq.

@@

Briefly, the raw deep sequencing data were processed using the 10X Genomics software package CellRanger (version 2.0.0).
The reads were aligned to a concatenation of the human and influenza virus transcriptomes.
The human transcriptome was generated by filtering genome assembly GRCh38 for protein coding genes defined in the GTF file GRCh38.87.
The influenza virus transcriptome was generated from the reverse-complement of the wildtype WSN viral RNA sequences as encoded in the reverse-genetics plasmids (Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}).
The aligned deep sequencing data are available on the GEO repository under accession GSE108041 (\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE108041}).

CellRanger calls cells based on the number of observed cell barcodes, and creates a cell-gene matrix.
We used custom Python code to annotate the cells in this matrix by the number of viral reads that could be assigned to the wildtype and synonymously barcoded virus.
Only about half of the viral reads overlapped the barcoded regions of the genes (Figure~\ref{fig:workflow}A) and could therefore be assigned to a viral barcode (Figure~\ref{fig:viralbarcodes}-Figures~supplement~\ref{figsupp:barcodescalled}).
So for calculations of the number of reads in a cell derived from each viral barcode for each viral gene, the total number of detected molecules of that gene are multiplied by the fraction of those molecules with assignable barcodes that are assigned to that barcode.
This annotated cell-gene matrix is in Supplementary~file~\ref{suppfile:cellgenematrix}.
A Jupyter notebook that performs these analyses is in Supplementary~file~\ref{suppfile:code}.
